Using the internal language of toposes in commutative algebra

### 2017-09-15 slides-pssl101.tex
Slides almost finished

@@

  PSSL 101 in Leeds \\
  \ \\
  September 17st, 2017
  \par
\end{frame}

@@

\section[Internal language]{The internal language of toposes}

@@

  \begin{block}{Examples}\begin{itemize}
    \item \hil{$\Set$:} \tabto{1.4cm} category of sets
    \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
    \item \hil{$\Zar(S)$:} \tabto{1.4cm} big Zariski topos of a base scheme~$S$
  \end{itemize}\end{block}

@@

\frame[t]{\frametitle{The little Zariski topos}
  Let~$A$ be a ring.
  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.
  \hil{Yes,} if we allow a change of topos:
  Then $A \to \O_{\Spec(A)}$ is the universal localisation.

  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}

  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
  for some~$g_i$.
}

% Wieso hat LocRing --> Ring keinen Linksadjungierten?
% Sollte also nicht Limiten bewahren.

% Recall the definition of a local ring and a local homomorphism!

\subsection{What is the internal language?}

\frame[t]{\frametitle{What is the internal language?}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

@@

\section[Little Zariski]{The little Zariski topos of a ring}

\begin{frame}{XXX}
\end{frame}

% O_{Spec(A)} as the reification of all stalks
% O_{Spec(A)} integral domain iff all stalks integral domains
% Transfer theorem

% *Constructed* to yield a universal localisation (mention classifying topos),
% *observed* to fulfil further properties.
% Give Mulvey/Tierney citation, state background (quasicoherence)
% also neg neg (f = 0) ==> f = 0

% Anonymous Noetherian condition
% Prove it

% Grothendieck's generic freeness lemma

% Outlook

\frame[t]{\frametitle{The little Zariski topos}
  Let~$A$ be a commutative ring (in~$\Set$).

  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.

  \hil{Yes,} if we allow a change of topos:
  Then $A \to \O_{\Spec(A)}$ is the universal localisation.

  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}

  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
  for some~$g_i$.

  \begin{itemize}
    \item Internally, the structure sheaf~$\O_X$ looks like \[ \text{
    an ordinary ring.} \]
    \item Internally, a sheaf of~$\O_X$-modules looks like \[
    \text{an ordinary module on that ring.} \]
  \end{itemize}
}

@@

\subsection[Unique features]{Unique features of the internal world}

\begin{frame}\frametitle{Unique features of the internal world}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

  \only<1>{
    \centering
    \includegraphics[scale=0.3]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
    \tiny
    %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
    %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
    %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
    %pp.~189--210.
    Miles Tierney. On the spectrum of a ringed topos. 1976.
    \par
  }
  \pause

  If~$X$ is reduced, this implies:
  \begin{itemize}
    \item $\O_X$ is a \hil{field} in that $\neg(\text{$x$ invertible})
    \Rightarrow x = 0$.
    \item $\O_X$ has \hil{$\neg\neg$-stable equality}: $\neg\neg(x = 0)
    \Rightarrow x = 0$.
    \item $\O_X$ is \hil{anonymously Noetherian}: Any ideal of~$\O_X$ is \mbox{\hil{not
    not}} finitely generated.
  \end{itemize}
\end{frame}

\subsection[Generic freeness]{Grothendieck's generic freeness lemma}

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  \hil{Lemma.} If $R$ is reduced, there
  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  is $f \neq 0$ in $R$ such that
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
  $\O_{\Spec R}$ is a \hil{field} and \hil{anonymously Noetherian}.
\end{frame}

\subsection{Transfer principles}

\begin{frame}\frametitle{Transfer principles}

  \hil{Question:} How do the properties of
  \begin{itemize}
  \item an~$A$-module~$M$ in~$\Set$ and
  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
  \end{itemize}
  \pause
  \vfill

  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
  \begin{itemize}
  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
  \end{itemize}
  Note: $M$ and $\underline{M}$ share all first-order properties.
  \pause
  \vfill

  \centering
  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
  \hil{stable under localisation}.
  \par
\end{frame}

>>>>

  \small
  ``Without loss of generality, any reduced ring is Noetherian and a field.''

@@

  101st Peripatetic Seminar on Sheaves and Logic (Leeds) \\
  September 17st, 2017
  \par
\end{frame}

@@

\section[Internal language of toposes]{The internal language of toposes}

@@

  \begin{block}{Examples}\begin{itemize}
    \item \hil{$\Set$:} \tabto{1.4cm} the category of sets
    \item \hil{$\Sh(X)$:} \tabto{1.4cm} the category of set-valued sheaves on a space~$X$
    \item \hil{$\Zar(S)$:} \tabto{1.4cm} the big Zariski topos of a base scheme~$S$
  \end{itemize}\end{block}

@@

\subsection{Universal localisation}
\frame[t]{\frametitle{Universal localisation}
  \setbeamercolor{block title}{bg=black!30,fg=black}
  \begin{block}{Recall}
    A ring is local iff it has precisely one maximal ideal.

    A homomorphism is local iff it reflects invertibility.
  \end{block}

  Let~$A$ be a ring.
  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.
  \hil{Yes,} if we allow a change of topos:
  Then $A \to A^\sim$ is the universal localisation.
}

\subsection{What is the internal language?}
\frame[t]{\frametitle{What is the internal language?}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

@@

\begin{frame}[c]\frametitle{Praise for Mike Shulman}
  \centering
  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
  \par
\end{frame}

\section[The little Zariski topos]{The little Zariski topos of a ring}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{spectrum~$\boldsymbol{\Spec(A)}$} is
  \begin{itemize}
    \item generated by opens~$D(f)$ for~$f \in A$
    \item subject to~$\Spec(A) = \bigcup_i D(f_i)$ iff~$1 = \sum_i g_i f_i$
    for some~$g_i$.
  \end{itemize}

  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
  with~$A^\sim(D(f)) = A[f^{-1}]$.

  \begin{block}{Motto}The structure sheaf~$A^\sim$ is a reification of
  all of the stalks~$A_\ppp$.\end{block}

  For instance, all stalks~$A_\ppp$ are integral domains if and only if \[ \Spec(A) \models
  \speak{$A^\sim$ is an integral domain}. \]
\end{frame}

\subsection{Transfer principles}
\begin{frame}{Transfer principles}
  \begin{block}{Theorem}
    The structure sheaf~$A^\sim$ inherits all first-order properties
    of~$A$ which are stable under localisation.
  \end{block}

  \hil{Proof.} The structure sheaf~$A^\sim$ is the localisation
  \[ \underline{A}[\F^{-1}] \]
  of the constant sheaf~$\underline{A}$ at the \hil{generic filter}~$\F$, a
  sheaf with
  \[ \F_\ppp = A \setminus \ppp. \]
  The rings~$A$ and~$\underline{A}$ share all first-order
  properties.
\end{frame}

\subsection[Unique features]{Unique features of the internal world}
\begin{frame}{Unique features of the internal world}
  Internally to~$\Sh(\Spec(A))$,
  \begin{center}
    \vspace*{-1em}
    \hil{any non-invertible element of~$\boldsymbol{A^\sim}$ is nilpotent.}
  \end{center}

  \only<1>{
    {
      \centering
      \includegraphics[scale=0.25]{images/tierney-on-the-spectrum-of-a-ringed-topos} \\
      \tiny
      %Miles Tierney. ``On the spectrum of a ringed topos''. \\ In: \emph{Algebra,
      %Topology, and Category Theory. A Collection of Papers in Honor of Samuel
      %Eilenberg}. Ed.\@ by A.~Heller and M.~Tierney. Academic Press, 1976,
      %pp.~189--210.
      Miles Tierney. On the spectrum of a ringed topos. 1976.
      \par
    }
    \bigskip

    \scriptsize\justifying
    A sheaf~$\E$ of~$A^\sim$-modules is quasicoherent if and only if,
    internally to~$\Sh(\Spec(A))$, the set $\E[f^{-1}]$ is a $\nabla_f$-sheaf for
    any~$f : A^\sim$, where~$\nabla_f\varphi \defeqv (\text{$f$
    invertible} \Rightarrow \varphi)$.
    \par
  }

  \only<2>{
    If~$A$ is reduced, then furthermore~$A^\sim$ is
    \begin{itemize}
      \item reduced,
      \item a \hil{field} in that non-invertible elements
      are zero,
      \item \hil{anonymously Noetherian} in that
      any ideal is \hil{not not} finitely generated,
      \item and has~\hil{$\boldsymbol{\neg\neg}$-stable equality}:
      $\neg\neg(f = 0) \Longrightarrow f = 0$
    \end{itemize}
  }
\end{frame}

\subsection[Generic freeness]{Grothendieck's generic freeness lemma}
\begin{frame}\frametitle{Generic freeness}
  Let~$A$ be a reduced ring and~$B, M$ as follows:

  \vspace*{-4.8em}
  ${\qquad\qquad\quad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}\xymatrixcolsep{3.5pc}\xymatrixrowsep{2pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    A \ar[r]_{\text{of finite type}} & B
  }$

  \hil{Theorem.} There exists $f \neq 0$ in $A$ such that
  \begin{enumerate}
    \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
    \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
  \end{enumerate}
  \pause

  \hil{Constructive version.} If zero is the only
  element~$f \in A$ such that
  {\renewcommand{\insertenumlabel}{1}\usebeamertemplate{enumerate item}},
  {\renewcommand{\insertenumlabel}{2}\usebeamertemplate{enumerate item}}, and
  {\renewcommand{\insertenumlabel}{3}\usebeamertemplate{enumerate item}},
  then~$1 = 0 \in A$.
  \pause

  \hil{Proof.} Internally in~$\Sh(\Spec(A))$,
  \begin{enumerate}
    \item $B^\sim$ and~$M^\sim$ are \hil{not not} free
    over~$A^\sim$
    \item $A^\sim \to B^\sim$ is \hil{not not} of finite presentation,
    and
    \item $M^\sim$ is \hil{not not} finitely presented as a module
    over~$B^\sim$.
  \end{enumerate}
\end{frame}

% *Constructed* to yield a universal localisation (mention classifying topos),
% *observed* to fulfil further properties.

@@
